Escape Analysis and Specialization in a Tracing JIT

### 2010-09-27 talk/pepm2011/paper.tex
Tweaks.

The tracing JIT approach that the PyPy project is taking removes the overhead
of bytecode dispatch. In this paper we want to explain how the traces that are
produced by PyPy's tracing JIT can be optimized to also remove some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. To understand the problem more closely, we analyze the
occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
technique to achieve this is a form of escape analysis \cite{XXX} that we call
\emph{virtual objects}, which is described in Section~\ref{sec:virtuals}. The
goal of virtual objects is to remove allocations of temporary objects that have
a predictable lifetime and to optimize type dispatching in the process.

@@

@@

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are indented to
correspond to the stack level of the function that contains the traced
operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new}, \texttt{set/getfield\_gc} and
\texttt{guard\_class} operations.

@@

Some of the operations within this trace are \texttt{new} operations, which each
create a new instance of some class. These instances are used for a while, e.g.
by calling methods on them (which are inlined into the trace), reading and
writing their fields. Some of these instances \emph{escape}, which means that
they are stored in some globally accessible place or are passed into a function.

@@

    \item Category 4: Objects that live for a while, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used\footnote{In theory, the
    approach of Section~\ref{sec:XXX} works also for objects that live for
    exactly $n>1$ iterations and then don't escape, but we expect this to be a
    very rare case, so we do not handle it.}
\end{itemize}

@@

The main insight to improve the code shown in the last section is that objects
in category 1 don't survive very long and are collected by the garbage collector
soon after their allocation. Moreover, they are used only inside the loop and
nobody else in the program stores a reference to them. The idea for improving
the code is thus to analyze which objects fall in category 1 and which may thus
not be allocated at all.

@@

This process is called \emph{escape analysis}. The escape analysis of
our tracing JIT works by using \emph{virtual objects}: The trace is walked from
beginning to end and whenever a \texttt{new} operation is seen, the operation is
removed and a virtual object is constructed. The virtual object summarizes the
shape of the object that is allocated at this position in the original trace,
and is used by the optimization to improve the trace. The shapes describe
where the values that would be stored in the fields of the allocated objects
come from. Whenever the optimizer sees a \texttt{setfield} that writes into a virtual
object, that shape summary is thus updated and the operation can be removed.
When the optimizer encounters a \texttt{getfield} from a virtual, the result is read
from the virtual object, and the operation is also removed.

@@

@@

@@

@@

Looking at the final trace it is also completely clear that specialization has
happened. The trace corresponds to the situation in which the trace was
originally recorded, which happened to be a loop where \texttt{BoxedIntegers} were
used. The now resulting loop does not refer to the \texttt{BoxedInteger} class at
all any more, but it still has the same behaviour. If the original loop had
used \texttt{BoxedFloats}, the final loop would use \texttt{float\_*} operations
everywhere instead (or even be very different, if the object model had
user-defined classes).

>>>>

The tracing JIT approach that the PyPy project is taking removes the overhead
of bytecode dispatch. In this paper we want to explain how the traces that are
produced by PyPy's tracing JIT can be optimized to also remove some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. To understand the problem more closely, we analyze the
occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
technique to achieve this is a form of escape analysis \cite{XXX} that we call
\emph{virtual objects},\footnote{The terminology comes from \cite{Psyco}}
which is described in Section~\ref{sec:virtuals}. The
goal of virtual objects is to remove allocations of temporary objects that have
a predictable lifetime and to optimize type dispatching in the process.

@@

XXX this is not an RPython interpreter; put a reference to the previous
paper to show how we deal with an interpreted piece of code and remove
the interpretation overhead, turning it into basically something
equivalent to the example here, which is the start of the present paper.

@@

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new}, \texttt{set/getfield\_gc} and
\texttt{guard\_class} operations.

@@

Some of the operations within this trace are \texttt{new} operations, which each
create a new instance of some class. These instances are used for a while, e.g.
by calling methods on them (which are inlined into the trace), reading and
writing their fields. Some of these instances \emph{escape}, which means that
they are stored in some globally accessible place or are passed into a
non-inlined function via a residual call.

@@

    \item Category 4: Objects that live for a while, survive across the jump,
    and then escape. To these we also count the objects that live across several
    jumps and then either escape or stop being used.\footnote{In theory, the
    approach of Section~\ref{sec:XXX} works also for objects that live for
    exactly $n>1$ iterations and then don't escape, but we expect this to be a
    very rare case, so we do not handle it.}
\end{itemize}

@@

The main insight to improve the code shown in the last section is that objects
in category 1 don't survive very long -- they are used only inside the loop and
nobody else in the program stores a reference to them. The idea for improving
the code is thus to analyze which objects fall in category 1 and may thus
not be allocated at all.

@@

This process is called \emph{escape analysis}. The escape analysis of
our tracing JIT works by using \emph{virtual objects}: The trace is walked from
beginning to end and whenever a \texttt{new} operation is seen, the operation is
removed and a virtual object\footnote{XXX what I have in mind when I talk
of ``virtual object'' is the run-time behavior -- i.e. a real object that
would exist at run-time, except that it has be virtual-ized.  Here you seem
to mean rather ``virtual object description'' or something.}
is constructed. The virtual object summarizes the
shape of the object that is allocated at this position in the original trace,
and is used by the optimization to improve the trace. The shapes describe
where the values that would be stored in the fields of the allocated objects
come from. Whenever the optimizer sees a \texttt{setfield} that writes into a virtual
object, that shape summary is thus updated and the operation can be removed.
When the optimizer encounters a \texttt{getfield} from a virtual, the result is read
from the virtual object, and the operation is also removed.

@@

XXX ``the trace above'' is dangerous; should mention all figures by numbers

@@

XXX should the variables be written in $math-style$ everywhere?

@@

XXX the figure is moved elsewhere by latex

@@

Looking at the final trace it is also completely clear that specialization has
happened. The trace corresponds to the situation in which the trace was
originally recorded, which happened to be a loop where \texttt{BoxedIntegers} were
used. The now resulting loop does not refer to the \texttt{BoxedInteger} class at
all any more, but it still has the same behaviour. If the original loop had
used \texttt{BoxedFloats}, the final loop would use \texttt{float\_*} operations
everywhere instead (or even be very different, if the object model had
more different classes).
